---
title: "Midterm Challenge: Philadelphia Housing Price Prediction"
subtitle: "MUSA 5080 - Public Policy Analytics"
author: "Vedika Jawa"
date: "March 16, 2026"
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## The Challenge

You are consultancy (please name your consultancy) competing to win the bid to work for a project for the **Philadelphia Office of Property Assessment**. The city wants to improve its Automated Valuation Model (AVM) for property tax assessments. Your task is to build a predictive model for residential sale prices and **present your findings** to city officials in a 5-minute briefing. 

**Deliverables:**

1. **Presentation slides** (10-15 slides MAX) - Your main findings for stakeholders
2. **Technical appendix** (HTML document) - All code, diagnostics, and detailed analysis
3. **5-minute in-class presentation** - Deliver your slides on October 27th

**Your goal:** Predict 2023-2024 home sale prices accurately while communicating findings clearly to a policy audience.

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**What drives the prices of residential properties to improve the Automated Valuation Model for property tax assessments in Philadelphia?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Load Libraries

```{r}
#Clearing previous data
rm(list=ls())

library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(lubridate)
library(janitor)
library(knitr)

options(tigris_use_cache = TRUE)

```

**Questions to answer:**

---

#### Step 2: Load & Inspect Philadelphia Property Sales

```{r}

sales_raw <- st_read("Data/opa_properties_public.geojson") %>%
  clean_names()

glimpse (sales_raw)
```

**Questions to answer:**

---

#### Step 3: Clean & Filter Sales Data

```{r}
# 3.1 Standardize & Parse Dates

sales_clean <- sales_raw %>%
  mutate(
    sale_date = as.Date(sale_date),
    year = year(sale_date)
  )

#3.2 Filter to Residential Properties (2023–2024)
#Decision: Keep only arms-length residential transactions between 2023–2024.

sales_clean <- sales_clean %>%
  filter(
    year >= 2023 & year <= 2024,
    category_code %in% c(1, 2, 3)
  ) 

# philly_sales_res <- philly_sales_2324 %>%
#   filter(
#     str_detect(category_code, 
#                "1|2|3")
#   ) %>%
#   # Remove vacant land explicitly
#   filter(!str_detect(category_code_description, "VACANT|SPECIAL PURPOSE"))

#3.3 Remove Obvious Data Errors
# Cleaning Decisions:
# Remove sale_price < $10,000 (likely transfer or recording errors)
# Remove sale_price > $5,000,000 (extreme luxury outliers)
# Remove missing coordinates
# Remove missing key structural attributes

sales_clean <- sales_clean %>%
  filter(
    !is.na(sale_price),
    sale_price > 10000
  ) %>%
  filter(!is.na(number_of_bedrooms),
         !is.na(number_of_bathrooms),
         !is.na(location))


# 3.4 Handle Missing Values
# Decision:
# Drop observations missing key modeling variables
# Do not impute sale price
# Impute interior square footage with median if missing

sales_clean <- sales_clean %>%
  mutate(
    total_livable_area = ifelse(
      is.na(total_livable_area),
      median(total_livable_area, na.rm = TRUE),
      total_livable_area
    )
  )

#3.5 Summary: Before vs After Cleaning

dim_after <- dim(sales_clean)
dim_before <- dim(sales_raw)
summary_table <- tibble(
  Stage = c("Raw OPA Data", "Cleaned Residential Sales 2023-2024"),
  Observations = c(dim_before[1], dim_after[1]),
  Variables = c(dim_before[2], dim_after[2])
)

kable(summary_table)

```

**Questions to answer:**

---

#### Step 4: Convert OPA Geometry to sf

```{r}

sales_sf <- st_as_sf(sales_clean, wkt = "shape", crs = 2272)

st_crs(sales_sf)

```

**Questions to answer:**

---

#### Step 5: Load Census Data (Block Group Level)

ACS 2023 5-year estimates
Geography: Block Group
County: Philadelphia

```{r}
# 5.1 Define Variables

acs_vars <- c(
  # Population
  total_population = "B01003_001",
  
  # Income
  median_income = "B19013_001",

  # Poverty
  poverty_count = "B17001_002",
  poverty_total = "B17001_001",

  # Education (Bachelor's Degree)
  bachelors = "B15003_022",
  education_total = "B15003_001",

  # Employment
  employed = "B23025_004",
  labor_force = "B23025_003",

  # Owner Occupancy
  owner_occupied = "B25003_002",
  housing_total = "B25003_001"
)

#5.2 Download Census Data
phl_acs <- get_acs(
  geography = "block group",
  variables = acs_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  output = "wide",
  geometry = TRUE
)

#5.3 Reshape & Create Derived Variables
# phl_acs_wide <- phl_acs %>%
#   select(GEOID, variable, estimate, geometry) %>%
#   st_drop_geometry() %>%
#   pivot_wider(names_from = variable, values_from = estimate)

phl_acs_wide <- phl_acs %>%
  st_drop_geometry()

phl_geometry <- phl_acs %>%
  select(GEOID, geometry) %>%
  distinct()

#5.4 Create Derived Socioeconomic Variables
phl_acs_sf <- left_join(phl_geometry, phl_acs_wide, by = "GEOID") %>%
  st_as_sf() %>%
  mutate(

    # Poverty rate
    poverty_rate = poverty_countE / poverty_totalE,

    # Percent bachelor degree holders
    pct_bachelors = bachelorsE / education_totalE,

    # Employment rate
    employment_rate = employedE / labor_forceE,

    # Owner occupancy rate
    owner_occ_rate = owner_occupiedE / housing_totalE
  )

#5.4 Match CRS
phl_acs_sf <- st_transform(phl_acs_sf, 2272)

```

**Questions to answer:**

---

#### Step 6: Load Spatial Amenities (OpenDataPhilly Example: Parks)
Assume parks shapefile downloaded from OpenDataPhilly.

```{r}

parks <- st_read("Data/PPR_Properties/PPR_Properties.shp") %>%
  st_transform(2272)

bus_stops <- st_read("Data/Fall_2022_Stop_Summary_(Bus)/Fall_2022_Stop_Summary_(Bus).shp") %>%
  st_transform(2272)

trolley_stops <- st_read("Data/Trolley_Stations/Trolley_Stations.shp") %>%
  st_transform(2272)

crime_data <- crime <- st_read("Data/crime data/incidents_part1_part2/incidents_part1_part2.shp") %>%
  st_transform(2272)

schools <- st_read("Data/PaSchoolDistricts2024_03/PaSchoolDistricts2024_03.shp") %>%
  st_transform(2272)

restaurants <- st_read("Data/NeighborhoodFoodRetail-shp/54b2029c-710c-40e7-8746-b90b123864752020329-1-yv09k6.t812l.shp") %>%
  st_transform(2272)

hospitals <- st_read("Data/Hospitals/Hospital_Data.shp") %>%
  st_transform(2272)

```

**Questions to answer:**

---
#### Step 7: External Data - Distance to Downtown Philadelphia

```{r}

#7.1 Define downtown centroid (Philadelphia City Hall)
downtown <- st_sfc(
  st_point(c(-75.16379, 39.95233)),
  crs = 4326
) %>%
  st_transform(2272)

#Compute centroid of each block group
phl_acs_sf <- phl_acs_sf %>%
  mutate(bg_centroid = st_centroid(geometry))

#7.2 Calculate Distance to Nearest Park
phl_acs_sf$dist_to_downtown <-
  st_distance(phl_acs_sf$bg_centroid, downtown)

```

**Questions to answer:**

---

#### Step 8: Spatial Feature engineering

```{r}

# 8.1 Create centroids for block groups
bg_centroids <- st_centroid(phl_acs_sf)

#8.2 Distance to Transit
# Combine transit layers
transit_stops <- bus_stops

# Distance to nearest transit stop
phl_acs_sf$distance_to_transit <-
  apply(st_distance(bg_centroids, transit_stops), 1, min)

#8.3 Distance to Parks
phl_acs_sf$distance_to_park <-
  apply(st_distance(bg_centroids, parks), 1, min)

#8.4 Distance to Hospitals
phl_acs_sf$hospital_distance <-
  apply(st_distance(bg_centroids, hospitals), 1, min)

#8.5 Crime Density
crime_join <- st_join(crime_data, phl_acs_sf, join = st_within)

crime_counts <- crime_join %>%
  st_drop_geometry() %>%
  count(GEOID, name = "crime_count")

#Join to census
phl_acs_sf <- phl_acs_sf %>%
  left_join(crime_counts, by = "GEOID")

#Replacing missing values
phl_acs_sf$crime_count[is.na(phl_acs_sf$crime_count)] <- 0

#Calculate Crime Density
phl_acs_sf$area_sqmi <- as.numeric(st_area(phl_acs_sf)) / 27878400

phl_acs_sf$crime_density <- phl_acs_sf$crime_count / phl_acs_sf$area_sqmi

#8.6 Restaurant Density
restaurant_join <- st_join(restaurants, phl_acs_sf, join = st_within)

restaurant_counts <- restaurant_join %>%
  st_drop_geometry() %>%
  count(GEOID, name = "restaurant_count")

phl_acs_sf <- phl_acs_sf %>%
  left_join(restaurant_counts, by = "GEOID")

phl_acs_sf$restaurant_count[is.na(phl_acs_sf$restaurant_count)] <- 0

phl_acs_sf$restaurant_density <-
  phl_acs_sf$restaurant_count / phl_acs_sf$area_sqmi

#8.7 School Density
school_join <- st_join(schools, phl_acs_sf, join = st_within)

school_counts <- school_join %>%
  st_drop_geometry() %>%
  count(GEOID, name = "school_count")

phl_acs_sf <- phl_acs_sf %>%
  left_join(school_counts, by = "GEOID")

phl_acs_sf$school_count[is.na(phl_acs_sf$school_count)] <- 0

phl_acs_sf$school_density <-
  phl_acs_sf$school_count / phl_acs_sf$area_sqmi

#8.8 Feature Check
phl_acs_sf %>%
  st_drop_geometry() %>%
  select(
    GEOID,
    distance_to_transit,
    distance_to_park,
    hospital_distance,
    crime_density,
    restaurant_density,
    school_density
  ) %>%
  summary()

```

**Questions to answer:**

---

#### Step 9: Final Dataset Summary
Assume parks shapefile downloaded from OpenDataPhilly.

```{r}

#Ataching block features to property sales
sales_sf <- st_transform(sales_sf, 2272)

sales_joined <- st_join(sales_sf, phl_acs_sf)

#Cross Checking if the table worked
final_dim <- dim(sales_joined)

final_summary <- tibble(
  Stage = c("Cleaned Sales", "Final Spatial Dataset"),
  Observations = c(dim_after[1], final_dim[1]),
  Variables = c(dim_after[2], final_dim[2])
)

kable(final_summary)

```

**Questions to answer:**

---

## Part 2: Exploratory Data Analysis

### Research Question

Create at least 5 professional visualizations: 
Distribution of sale prices (histogram) 
Geographic distribution (map) 
Price vs. structural features (scatter plots) 
Price vs. spatial features (scatter plots) 
One creative visualization

#### Step 1: Distribution of Sale Prices (Histogram)
Assume parks shapefile downloaded from OpenDataPhilly.

```{r}

library(tidyverse)
library(sf)
library(scales)
library(viridis)
library(janitor)
# Create a plotting dataset
sales_plot <- sales_joined %>%
  mutate(
    log_sale_price = log(sale_price),
    total_livable_area = as.numeric(total_livable_area),
    number_of_bedrooms = as.numeric(number_of_bedrooms),
    number_of_bathrooms = as.numeric(number_of_bathrooms),
    distance_to_transit = as.numeric(distance_to_transit),
    distance_to_park = as.numeric(distance_to_park),
    hospital_distance = as.numeric(hospital_distance),
    crime_density = as.numeric(crime_density),
    restaurant_density = as.numeric(restaurant_density),
    dist_to_downtown = as.numeric(dist_to_downtown)
  )

```
```{r}

ggplot(st_drop_geometry(sales_plot), aes(x = log_sale_price)) +
  geom_histogram(bins = 50, fill = "#2C7FB8", color = "white") +
  labs(
    title = "Distribution of Philadelphia Residential Sale Prices",
    subtitle = "Log transformation reduces extreme skew",
    x = "Log Sale Price",
    y = "Number of Sales"
  ) +
  theme_minimal(base_size = 13)
```
```{r}
bg_price_map <- sales_plot %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(
    median_sale_price = median(sale_price, na.rm = TRUE),
    sales_count = n()
  )

bg_map <- phl_acs_sf %>%
  left_join(bg_price_map, by = "GEOID")

ggplot(bg_map) +
  geom_sf(aes(fill = median_sale_price), color = NA) +
  scale_fill_viridis_c(
    option = "magma",
    labels = label_dollar(),
    na.value = "grey90"
  ) +
  labs(
    title = "Median Home Sale Price by Census Block Group",
    subtitle = "Philadelphia residential sales, 2023–2024",
    fill = "Median Price"
  ) +
  theme_minimal(base_size = 13)
```
```{r}
ggplot(
  st_drop_geometry(sales_plot) %>%
    filter(!is.na(total_livable_area), total_livable_area > 0, total_livable_area < 6000),
  aes(x = total_livable_area, y = sale_price)
) +
  geom_point(alpha = 0.15, size = 1, color = "#3182BD") +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Larger Homes Sell for More",
    subtitle = "Sale price rises with total livable area",
    x = "Total Livable Area (sq ft)",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
ggplot(
  st_drop_geometry(sales_plot) %>%
    filter(!is.na(number_of_bathrooms), number_of_bathrooms <= 6),
  aes(x = factor(number_of_bathrooms), y = sale_price)
) +
  geom_boxplot(fill = "#9ECAE1", outlier.alpha = 0.15) +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Homes With More Bathrooms Sell for More",
    subtitle = "Distribution of sale prices by number of bathrooms",
    x = "Number of Bathrooms",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 13)
```
```{r}
ggplot(
  st_drop_geometry(sales_plot),
  aes(x = dist_to_downtown, y = log_sale_price)
) +
  geom_point(alpha = 0.15, color = "#756BB1") +
  geom_smooth(method = "loess", se = FALSE, color = "black") +
  labs(
    title = "Homes Closer to Downtown Tend to Sell for More",
    x = "Distance to Downtown (feet)",
    y = "Log Sale Price"
  ) +
  coord_cartesian(xlim = c(0, 40000)) +
  theme_minimal()
```
```{r}
zip_summary <- sales_plot %>%
  st_drop_geometry() %>%
  group_by(zip_code) %>%
  summarize(
    median_sale_price = median(sale_price, na.rm = TRUE),
    sales_count = n()
  ) %>%
  filter(!is.na(zip_code), sales_count >= 50) %>%
  arrange(desc(median_sale_price)) %>%
  slice_head(n = 10)

ggplot(zip_summary, aes(x = reorder(zip_code, median_sale_price), y = median_sale_price)) +
  geom_col(fill = "#DD3497") +
  coord_flip() +
  scale_y_continuous(labels = label_dollar()) +
  labs(
    title = "Top 10 ZIP Codes by Median Sale Price",
    subtitle = "Among ZIP codes with at least 50 sales",
    x = "ZIP Code",
    y = "Median Sale Price"
  ) +
  theme_minimal(base_size = 13)
```

**Questions to answer:**

---

#### Step 2: Geographic Distribution (Map)
Assume parks shapefile downloaded from OpenDataPhilly.

```{r}

ggplot(sales_joined) +
  geom_sf(aes(color = sale_price), size = 0.8, alpha = 0.7) +
  scale_color_viridis(
    option = "plasma",
    labels = dollar_format()
  ) +
  labs(
    title = "Geographic Distribution of Sale Prices",
    color = "Sale Price"
  ) +
  theme_minimal(base_size = 14)

ggplot(sales_joined) +
  geom_sf(aes(color = log(sale_price)), size = 0.8, alpha = 0.7) +
  scale_color_viridis(option = "magma") +
  labs(
    title = "Geographic Distribution of Log Sale Prices",
    color = "Log Price"
  ) +
  theme_minimal(base_size = 14)



```

**Questions to answer:**

---

#### Step 3: Price vs Structural Features (Scatter Plots)
Assume parks shapefile downloaded from OpenDataPhilly.

```{r}
#A. Price vs Living Area
ggplot(sales_joined, aes(x = total_livable_area, y = sale_price)) +
  geom_point(alpha = 0.4, color = "#34495E") +
  geom_smooth(method = "lm", color = "red") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Sale Price vs Living Area",
    x = "Total Livable Area (sq ft)",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 14)

#B. Price vs Year Built
ggplot(sales_joined, aes(x = year_built, y = sale_price)) +
  geom_point(alpha = 0.4, color = "#7F8C8D") +
  geom_smooth(method = "loess", color = "blue") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Sale Price vs Year Built",
    x = "Year Built",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 14)


```

**Questions to answer:**

---

#### Step 4:Price vs Spatial Features

Assumes you created:
median_income
poverty_rate
dist_to_park

```{r}
#A. Price vs Neighborhood Income

ggplot(sales_joined, aes(x = median_income, y = sale_price)) +
  geom_point(alpha = 0.4, color = "#16A085") +
  geom_smooth(method = "lm", color = "black") +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Sale Price vs Neighborhood Median Income",
    x = "Median Household Income",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 14)


#B. Price vs Distance to Park
ggplot(sales_joined, aes(x = dist_to_park, y = sale_price)) +
  geom_point(alpha = 0.4, color = "#27AE60") +
  geom_smooth(method = "lm", color = "black") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Sale Price vs Distance to Nearest Park",
    x = "Distance to Park (feet)",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 14)

```

**Questions to answer:**

---

#### Step 5: Creative Visualization (Professional-Level)

Price per Square Foot Map
This is strong analytically.

```{r}
# This visual:
# Controls for house size
# Reveals neighborhood value patterns
# Looks very professional

sales_joined <- sales_joined %>%
  mutate(price_per_sqft = sale_price / total_livable_area)

ggplot(sales_joined) +
  geom_sf(aes(color = price_per_sqft), size = 0.8, alpha = 0.8) +
  scale_color_viridis(
    option = "inferno",
    labels = dollar_format()
  ) +
  labs(
    title = "Price Per Square Foot Across Philadelphia",
    color = "Price / Sq Ft"
  ) +
  theme_minimal(base_size = 14)

#Optional: Boxplot by Bedroom Count
ggplot(sales_joined, aes(x = factor(number_of_bedrooms), y = sale_price)) +
  geom_boxplot(fill = "#3498DB") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Sale Price by Number of Bedrooms",
    x = "Bedrooms",
    y = "Sale Price"
  ) +
  theme_minimal(base_size = 14)

```

**Questions to answer:**

---

## Part 4: Model Building

#### Step 4.2: Spatial Model

```{r}
#Prepare Modeling Dataset

model_spatial <- sales_joined %>%
  st_drop_geometry() %>%
  select(
    sale_price,
    distance_to_transit,
    distance_to_park,
    hospital_distance,
    crime_density,
    restaurant_density,
    school_density,
    dist_to_downtown
  ) %>%
  drop_na()

nrow(model_spatial)

#Transform Price (Log) as the Housing prices are right-skewed
model_spatial <- model_spatial %>%
  mutate(
    log_price = log(sale_price)
  )

#Spatial Regression Model
spatial_model <- lm(
  log_price ~
    distance_to_transit +
    distance_to_park +
    hospital_distance +
    crime_density +
    restaurant_density +
    school_density +
    dist_to_downtown,
  data = model_spatial
)

#Summary
summary(spatial_model)

#Standardizing Variables
model_spatial_scaled <- model_spatial %>%
  mutate(across(
    c(
      distance_to_transit,
      distance_to_park,
      hospital_distance,
      crime_density,
      restaurant_density,
      school_density,
      dist_to_downtown
    ),
    scale
  ))

#Visualisation
library(ggplot2)
library(units)

ggplot(spatial_model, aes(x = dist_to_downtown, y = log_price)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal() +
  labs(
    title = "Housing Prices vs Distance to Downtown",
    x = "Distance to Downtown (feet)",
    y = "Log Sale Price"
  )

```

**Questions to answer:**

This regression model examines how **spatial accessibility and neighborhood amenities** influence housing prices in Philadelphia. The dependent variable is **log-transformed sale price**, meaning the coefficients represent **approximate percentage changes in housing prices**.

The model is statistically significant overall (F-statistic p < 0.001), indicating that the spatial variables collectively help explain variation in property prices. However, the **Adjusted R² is 0.040**, meaning spatial features alone explain about **4% of the variation in housing prices**. This suggests that while location characteristics matter, **structural housing attributes and neighborhood socioeconomic conditions likely play a larger role in determining prices**.

Interpretation of Individual Variables

**Distance to Transit**

Coefficient: **3.58e-05**, p = **0.0047**

This variable is statistically significant. Surprisingly, the positive coefficient suggests that properties located **farther from transit stops are slightly more expensive**. This may reflect neighborhood patterns where higher-value residential areas are located farther from major transit corridors.

**Distance to Park**

Coefficient: **−1.35e-05**, p = **0.099**

The relationship is negative, meaning **homes closer to parks tend to have slightly higher prices**, but the effect is **only marginally significant** at the 10% level. This suggests that proximity to green space may modestly increase property values.

**Distance to Hospital**

Coefficient: **−2.83e-05**, p < **0.001**

This variable is highly significant. The negative coefficient indicates that **homes located closer to hospitals tend to have higher sale prices**. Hospitals may act as employment centers or be located in well-developed urban areas with higher housing demand.


**Crime Density**

Coefficient: **−4.20e-04**, p < **0.001**

Crime density has a **strong negative relationship with housing prices**. Areas with higher crime levels are associated with **lower property values**, which aligns with expectations in urban housing markets.


**Restaurant Density**

Coefficient: **−0.108**, p < **0.001**

This variable is statistically significant but negative, suggesting that **areas with higher concentrations of restaurants may have slightly lower housing prices** in this model. This could occur if restaurant-heavy areas are located in dense commercial districts or mixed-use neighborhoods where residential demand differs.


**Distance to Downtown**

Coefficient: **−2.97e-06**, p < **0.001**

The negative coefficient indicates that **homes located closer to downtown Philadelphia tend to have higher prices**, reflecting the value of central-city accessibility.


**School Density**

This variable was **dropped from the model due to singularity**, meaning it is perfectly correlated with one or more other predictors. This often occurs when spatial variables are highly correlated, such as school density and population density.


**Overall Interpretation**

The results indicate that **spatial accessibility and neighborhood amenities have statistically significant relationships with housing prices**, particularly crime density and proximity to downtown. However, the relatively low R² suggests that spatial features alone explain only a small portion of price variation.

This outcome is expected in housing market analysis, where **property-specific characteristics (size, bedrooms, age) and socioeconomic neighborhood conditions typically have stronger explanatory power than spatial accessibility alone**.

---

## Part 5: Model Evaluation

#### Step 5.1: Spatial Model

```{r}
install.packages("tidymodels")
library(tidymodels)

#Spatial Model
model_spatial <- sales_joined %>%
  st_drop_geometry() %>%
  select(
    sale_price,
    distance_to_transit,
    distance_to_park,
    hospital_distance,
    crime_density,
    restaurant_density,
    dist_to_downtown
  ) %>%
  mutate(
    distance_to_transit = as.numeric(distance_to_transit),
    distance_to_park = as.numeric(distance_to_park),
    hospital_distance = as.numeric(hospital_distance),
    dist_to_downtown = as.numeric(dist_to_downtown)
  ) %>%
  drop_na()

nrow(model_spatial)

colnames(model_spatial)

model_spatial <- model_spatial %>%
  mutate(log_price = log(sale_price))

colnames(model_spatial)

#10 fold cross validation

set.seed(123)

cv_folds <- vfold_cv(model_spatial, v = 10)

#Defining the Model
lm_model <- linear_reg() %>%
  set_engine("lm")

#Model Formula
spatial_formula <- log_price ~
  distance_to_transit +
  distance_to_park +
  hospital_distance +
  crime_density +
  restaurant_density +
  dist_to_downtown

#Run workflow
spatial_workflow <- workflow() %>%
  add_model(lm_model) %>%
  add_formula(spatial_formula)

#Run Cross Validation
control <- control_resamples(save_pred = TRUE)

cv_results <- fit_resamples(
  spatial_workflow,
  resamples = cv_folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control
)

#Collect Metrics
collect_metrics(cv_results)

#Predictions
cv_predictions <- collect_predictions(cv_results)

#Visualisation of Predicted vs Actual
ggplot(cv_predictions, aes(x = log_price, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal() +
  labs(
    title = "Predicted vs Actual Housing Prices",
    x = "Actual Log Sale Price",
    y = "Predicted Log Sale Price"
  )


```

**Questions to answer:**


---
## Part 6: Model Diagnostics

#### Step 6.1: Spatial Model

```{r}
spatial_model <- lm(
  log_price ~
    distance_to_transit +
    distance_to_park +
    hospital_distance +
    crime_density +
    restaurant_density +
    dist_to_downtown,
  data = model_spatial
)

#Residual Plot
library(ggplot2)

ggplot(data.frame(
  fitted = fitted(spatial_model),
  residuals = residuals(spatial_model)
), aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.3) +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal() +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  )

#QQ Plot

ggplot(data.frame(resid = residuals(spatial_model)), aes(sample = resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal() +
  labs(title = "Q-Q Plot of Spatial Model Residuals")

#Cooks Distance
#Calculate Cook’s Distance
cook <- cooks.distance(spatial_model)

#Convert to a Data Frame for ggplot
cook_df <- data.frame(
  observation = 1:length(cook),
  cook_distance = cook
)

#Cook’s Distance Plot 

ggplot(cook_df, aes(x = observation, y = cook_distance)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 4/nrow(cook_df), color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Cook's Distance for Spatial Model",
    x = "Observation",
    y = "Cook's Distance"
  )

which(cook > 4/nrow(cook_df))

sum(cook > 4/nrow(cook_df))

```


```{r}
## Model 4
full_model_data <- sales_joined %>%
  st_drop_geometry() %>%
  select(
    sale_price,

    # Structural variables
    total_livable_area,
    number_of_bedrooms,
    number_of_bathrooms,
    year_built,

    # Census variables
    median_incomeE,
    pct_bachelors,
    employment_rate,
    owner_occ_rate,
    poverty_rate,

    # Spatial variables
    distance_to_transit,
    distance_to_park,
    hospital_distance,
    crime_density,
    restaurant_density,
    dist_to_downtown,

    # Fixed effects variable
    census_tract
  ) %>%
  mutate(
    log_price = log(sale_price),

    # make sure numeric variables are truly numeric
    total_livable_area   = as.numeric(total_livable_area),
    number_of_bedrooms   = as.numeric(number_of_bedrooms),
    number_of_bathrooms  = as.numeric(number_of_bathrooms),
    year_built           = as.numeric(year_built),
    median_incomeE       = as.numeric(median_incomeE),
    pct_bachelors        = as.numeric(pct_bachelors),
    employment_rate      = as.numeric(employment_rate),
    owner_occ_rate       = as.numeric(owner_occ_rate),
    poverty_rate         = as.numeric(poverty_rate),
    distance_to_transit  = as.numeric(distance_to_transit),
    distance_to_park     = as.numeric(distance_to_park),
    hospital_distance    = as.numeric(hospital_distance),
    crime_density        = as.numeric(crime_density),
    restaurant_density   = as.numeric(restaurant_density),
    dist_to_downtown     = as.numeric(dist_to_downtown),
    census_tract         = as.factor(census_tract)
  ) %>%
  drop_na()

nrow(full_model_data)
glimpse(full_model_data)
```

```{r}
full_model_data <- sales_joined %>%
  st_drop_geometry() %>%
  select(
    sale_price,
    total_livable_area,
    number_of_bedrooms,
    number_of_bathrooms,
    median_incomeE,
    pct_bachelors,
    employment_rate,
    owner_occ_rate,
    distance_to_transit,
    distance_to_park,
    hospital_distance,
    crime_density,
    restaurant_density,
    dist_to_downtown,
    census_tract
  ) %>%
  mutate(
    log_price = log(sale_price),
    total_livable_area  = as.numeric(total_livable_area),
    number_of_bedrooms  = as.numeric(number_of_bedrooms),
    number_of_bathrooms = as.numeric(number_of_bathrooms),
    median_incomeE      = as.numeric(median_incomeE),
    pct_bachelors       = as.numeric(pct_bachelors),
    employment_rate     = as.numeric(employment_rate),
    owner_occ_rate      = as.numeric(owner_occ_rate),
    distance_to_transit = as.numeric(distance_to_transit),
    distance_to_park    = as.numeric(distance_to_park),
    hospital_distance   = as.numeric(hospital_distance),
    crime_density       = as.numeric(crime_density),
    restaurant_density  = as.numeric(restaurant_density),
    dist_to_downtown    = as.numeric(dist_to_downtown),
    census_tract        = as.character(census_tract)
  ) %>%
  filter(
    !is.na(log_price),
    !is.na(total_livable_area),
    !is.na(number_of_bedrooms),
    !is.na(number_of_bathrooms),
    !is.na(median_incomeE),
    !is.na(pct_bachelors),
    !is.na(employment_rate),
    !is.na(owner_occ_rate),
    !is.na(distance_to_transit),
    !is.na(distance_to_park),
    !is.na(hospital_distance),
    !is.na(crime_density),
    !is.na(restaurant_density),
    !is.na(dist_to_downtown),
    !is.na(census_tract)
  ) %>%
  mutate(
    census_tract = droplevels(factor(census_tract))
  )

full_model <- lm(
  log_price ~
    total_livable_area +
    number_of_bedrooms +
    number_of_bathrooms +
    median_incomeE +
    pct_bachelors +
    employment_rate +
    owner_occ_rate +
    distance_to_transit +
    distance_to_park +
    hospital_distance +
    crime_density +
    restaurant_density +
    dist_to_downtown +
    census_tract +
    median_incomeE:dist_to_downtown,
  data = full_model_data
)

summary(full_model)
```

```{r}
library(modelsummary)

modelsummary(
  full_model,
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```

```{r}
library(tidymodels)

set.seed(123)

full_folds <- vfold_cv(full_model_data, v = 10)

lm_spec <- linear_reg() %>%
  set_engine("lm")

full_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(
    log_price ~
      total_livable_area +
      number_of_bedrooms +
      number_of_bathrooms +
      median_incomeE +
      pct_bachelors +
      employment_rate +
      owner_occ_rate +
      distance_to_transit +
      distance_to_park +
      hospital_distance +
      crime_density +
      restaurant_density +
      dist_to_downtown +
      census_tract +
      median_incomeE:dist_to_downtown
  )

full_cv_results <- fit_resamples(
  full_workflow,
  resamples = full_folds,
  metrics = metric_set(rmse, mae, rsq),
  control = control_resamples(save_pred = TRUE)
)

collect_metrics(full_cv_results)
```


```{r}
full_cv_predictions <- collect_predictions(full_cv_results)

ggplot(full_cv_predictions, aes(x = log_price, y = .pred)) +
  geom_point(alpha = 0.2, color = "#2C7FB8") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Predicted vs Actual Log Sale Prices",
    x = "Actual Log Sale Price",
    y = "Predicted Log Sale Price"
  )
```


```{r}
# Residual plot
ggplot(
  data.frame(
    fitted = fitted(full_model),
    residuals = residuals(full_model)
  ),
  aes(x = fitted, y = residuals)
) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 0, color = "red") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  )

# QQ plot
ggplot(
  data.frame(resid = residuals(full_model)),
  aes(sample = resid)
) +
  stat_qq() +
  stat_qq_line(color = "red") +
  theme_minimal(base_size = 13) +
  labs(title = "Q-Q Plot of Full Model Residuals")

# Cook's distance
cook_full <- cooks.distance(full_model)

cook_full_df <- data.frame(
  observation = 1:length(cook_full),
  cook_distance = cook_full
)

ggplot(cook_full_df, aes(x = observation, y = cook_distance)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 4 / nrow(cook_full_df), color = "red", linetype = "dashed") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Cook's Distance for Full Model",
    x = "Observation",
    y = "Cook's Distance"
  )

sum(cook_full > 4 / nrow(cook_full_df))
```

**Questions to answer:**
Cook’s distance was used to identify influential observations in the spatial regression model. A threshold of 4/n.

4/n was used to flag potentially influential points. The Cook’s distance plot shows that only a small number of observations exceed this threshold, suggesting that the model estimates are not overly influenced by individual observations.

---